Peer influences in air travel behavior: A social autoregressive analysis

### 2015-02-18 article.Rmd
Fixed formatting of lists and added real citations to econometrics section.

Spatial Econometrics
---------------------------------------
Spatial econometrics is built on the understanding that observations near to
each other spatially are likely to exhibit correlation and interdependence, as
they share common unobserved variables and influence each other. A class of
models developed to solve these problems was pioneered by Anselin (1988);
LeSage & Pace (2009) provide a full and modern treatment.

For a linear regression model , a necessary assumption is that the observations
are independent, which is unlikely in spatial data. Tobler’s law implies that if
two observations are “close” to each other, then they are likely to share common
characteristics (Tobler 1970). Further, observations close to each other may
actually exert an influence on each other in the way air pollution lowers the
price of nearby properties (Kim & Goldsmith 2008). If correlation and
interdependence are not controlled in an econometric model, then the model may
return biased parameters and/or invalid hypothesis tests.

This correlation and interdependence can be accommodated with a spatial Durbin
model (SDM),

where  is a spatial weights matrix in which each element  maps the spatial
relationship between two observations  and ; if  and  are “neighbors,” then .
The parameter  measures the dependence between  and , and the lagged independent
variable coefficients in  contain the impact that attributes of  have on .

Such spatial models have been applied to numerous different contexts, including
transportation modeling and forecasting. In most if not all of these cases,
however, spatial data is used as a proxy for an inferred social relationship.
For example, Adjemian et al. (2010) observed that vehicle body types were
spatially autocorrelated, with the authors inferring that neighbors may
influence each other in choosing vehicle types. Similarly, Bhat et al.
(2010) used spatial proximity as a means to study the social destinations of
teenagers. In these and other cases, it may be more appropriate to use the
social network directly, if the social network can be translated to a  matrix.
This is effectively the approach of Bramoullé et al. (2009) and
Calvó-Armengol et al, (2009), who showed that a student's school performance
and physical activity level (respectively) are socially interdependent: students
with studious or athletic friends tend to be so as well.

There are two primary difficulties in creating , one econometric and one
practical. From an econometric standpoint, the researcher typically must specify
exogenously. Though some common practices have emerged in the spatial literature
(Dubin 1998), not all of these can be readily adopted to social networks.
This gap in the literature is one that we seek to fill with this study, by
examining several possible  matrices that we construct from social network data.
The practical difficulty is actually collecting that data, though this is
becoming less difficult as interaction within social networks happens through
online sites such as Facebook. We discuss this practical evolution in Section
2.2.

>>>>

Spatial Econometrics
---------------------------------------
Spatial econometrics is built on the understanding that observations near to
each other spatially are likely to exhibit correlation and interdependence, as
they share common unobserved variables and influence each other. A class of
models developed to solve these problems was pioneered by @Anselin1988;
@LeSage2009 provide a full and modern treatment.

For a linear regression model , a necessary assumption is that the observations
are independent, which is unlikely in spatial data. Tobler’s law implies that if
two observations are “close” to each other, then they are likely to share common
characteristics [@Tobler1970]. Further, observations close to each other may
actually exert an influence on each other in the way air pollution lowers the
price of nearby properties [@Kim2008]. If correlation and
interdependence are not controlled in an econometric model, then the model may
return biased parameters and/or invalid hypothesis tests.

This correlation and interdependence can be accommodated with a spatial Durbin
model (SDM),
$$y = \rho W y + X\beta + WX\gamma + \epsilon$$
where $W$ is a spatial weights matrix in which each element  maps the spatial
relationship between two observations $i$ and $j$; if $i$ and $j$ are
"neighbors," then $w_{ij} > 0$. The parameter $\rho$ measures the dependence
between $y_i$ and $y_j$, and the lagged independent variable coefficients in
$\gamma$ contain the impact that attributes of $j$ have on $y_i$.

Such spatial models have been applied to numerous different contexts, including
transportation modeling and forecasting. In most if not all of these cases,
however, spatial data is used as a proxy for an inferred social relationship.
For example, @Adjemian2010 observed that vehicle body types were
spatially autocorrelated, with the authors inferring that neighbors may
influence each other in choosing vehicle types. Similarly, @Bhat2010b used
spatial proximity as a means to study the social destinations of
teenagers, assuming that teenagers who live more closely to each other are more
likely to be in each other's social networks. In these and other cases, it may
be more appropriate to use the social network directly, if the social network
can be translated to a $W$ matrix. This is effectively the approach of
@Bramoulle2009 and @Calvo-Armengol2009, who showed that a student's school
performance and physical activity level (respectively) are socially
interdependent: students with studious or athletic friends tend to be so as
well.

There are two primary difficulties in creating , one econometric and one
practical. From an econometric standpoint, the researcher typically must specify
exogenously. Though some common practices have emerged in the spatial literature
[@Dubin1998], not all of these can be readily adopted to social networks.
This gap in the literature is one that we seek to fill with this study, by
examining several possible  matrices that we construct from social network data.
The practical difficulty is actually collecting that data, though this is
becoming less difficult as interaction within social networks happens through
online sites such as Facebook. We discuss this practical evolution in Section
2.2.
